---
title: "Writing controlledburn output in parallel"
description: "The serial COG write loop from the last post, replaced by 24 workers writing disjoint Zarr chunks. The 31-billion-cell grid on disk in 28 seconds of 9.5 MB."
author: "Michael Sumner"
date: "2026-09-03"
categories: [controlledburn, rasterization, Antarctica, GDAL, Zarr, mirai]
format:
  html:
    code-fold: false
    toc: true
execute:
  eval: false
---

In [Rasterizing ice-free Antarctica](../2026-08-19_antarctica-icefree-rasterizing/) the
rasterization itself was fast, and the slow part was the file. The
write loop visited about 8000 super-tiles, materializing each one from the sparse
burn and pushing it into a COG through a single dataset handle. That handle is the
whole problem: the COG driver's create path writes a temporary GeoTIFF and then
rewrites it on close, and neither step tolerates a second writer.

Every iteration of that loop is independent - crop, materialize, write - and the
burn object it reads from is 27 MB. The only thing shared is the sink. So change the
sink.

## Why Zarr

A Zarr array is a small metadata file plus one file per chunk. Two processes writing
different chunks never touch the same bytes, so there is nothing to coordinate. GDAL's
Zarr driver also declines to write a chunk that is entirely fill value (it will even
unlink an existing one), which means the store on disk *is* the sparse representation:
chunks over open ocean and interior ice sheet do not exist.

The plan is

1. create the empty store once, in the parent, with GDAL
2. open it in update mode from each worker, with its own handle, and write whole chunks
3. if a COG is still wanted, make it with one `translate()` at the end


## Instantiate

```{r}
library(controlledburn)
library(gdalraster)
library(mirai)

ext <- ba$extent
dm  <- ba$dimension
chunk_size <- 2048L
zarr_path  <- file.path(tempdir(), "rema_approx.zarr")

ds <- create(
  format = "Zarr", dst_filename = zarr_path,
  xsize = dm[1], ysize = dm[2], nbands = 1, dataType = "Int32",
  options = c(paste0("BLOCKSIZE=", chunk_size, ",", chunk_size),
              "COMPRESS=ZSTD", "FORMAT=ZARR_V2"),
  return_obj = TRUE)
ds$setGeoTransform(vaster::extent_dim_to_gt(ext, dm))
ds$setProjection(raster_info$projection$Wkt)
ds$setNoDataValue(band = 1, 0)   # fill_value; all-zero chunks are never written
ds$close()
```

The 4x4 "super-tile" from the last post is gone. A chunk write is one file write, so
the chunk is the natural unit of parallel work; 2048 keeps each dense allocation at
16 MB for Int32. (If 512-pixel chunks are wanted for cloud reads, set `BLOCKSIZE=512,512`
and keep iterating over 2048 windows; GDAL splits each write into 16 chunk files that
are still disjoint per worker.)

## Populate

The tile index and the pre-filter to rows that carry data are unchanged. The worker
opens the store in update mode, writes a batch of chunks, and closes. It calls nothing
that dirties metadata, so `.zarray` and `.zmetadata` are not rewritten on close.

```{r}
write_tiles <- function(tile_batch, zarr_path) {
  ds <- methods::new(gdalraster::GDALRaster, zarr_path, read_only = FALSE)
  on.exit(ds$close(), add = TRUE)
  n <- 0L
  for (i in seq_len(nrow(tile_batch))) {
    t <- tile_batch[i, ]
    sub <- controlledburn::crop_burn(ba, c(t$xmin, t$xmax, t$ymin, t$ymax))
    if (nrow(sub$runs) == 0 && nrow(sub$edges) == 0) next
    mat <- controlledburn::materialize_chunk(sub, fun = "id")
    mat[is.na(mat)] <- 0L
    ds$write(band = 1, xoff = t$offset_x, yoff = t$offset_y,
             xsize = t$ncol, ysize = t$nrow, rasterData = as.integer(t(mat)))
    n <- n + 1L
  }
  n
}
environment(write_tiles) <- globalenv()

daemons(24)
everywhere(library(controlledburn), ba = ba)   # ships ba once per daemon

batches <- split(tiles, cut(seq_len(nrow(tiles)), 24 * 4L, labels = FALSE))
system.time({
  res <- mirai_map(batches, write_tiles, .args = list(zarr_path = zarr_path))[.progress]
})
daemons(0)
```

Two `mirai` details that cost me a round trip. Objects passed through `...` to
`everywhere()` are assigned into each daemon's global environment, which is exactly
the "send the 27 MB once" behaviour wanted here; a plain `<-` inside the expression is
evaluated in a clean environment and lost (use `<<-` if you assign there). And
reparenting the worker function to `globalenv()` stops it dragging the Quarto chunk
environment along with every task.

## Result

On a 32-cpu, 64 GB machine with 24 daemons:

```
   user  system elapsed
  ...     ...    28.0
```

```{r}
library(terra)
rast(zarr_path)
#> class       : SpatRaster
#> size        : 182568, 170318, 1  (nrow, ncol, nlyr)
#> resolution  : 32, 32  (x, y)
#> extent      : -2700096, 2750080, -2500096, 3342080  (xmin, xmax, ymin, ymax)
#> source      : rema_approx.zarr

f0 <- fs::dir_ls(zarr_path, recurse = TRUE)
length(f0)
#> [1] 722
sum(fs::file_size(f0))
#> 9.48M
```

The grid could hold ceiling(170318 / 2048) x ceiling(182568 / 2048) = 84 x 90 = 7560
chunks. About 718 of them exist. That is 9.5% of the continent's bounding box with any
rock in it at the 2048-pixel scale, a number obtained by listing a directory, and the
whole store is 9.5 MB - alongside the 27 MB burn object, the 5.6 MB DEFLATE COG, and
the 613 MB uncompressed AADC product from last time, it is another point on the same
continuum.

## Coda: the COG is now cheap

If the COG is still the deliverable, the expensive part - materializing every tile -
has already happened in parallel. What remains is one sequential translate, I/O-bound,
with the COG driver's own threads handling compression and overviews:

```{r}
set_config_option("GDAL_NUM_THREADS", "ALL_CPUS")
translate(zarr_path, file.path(tempdir(), "rema_approx.tif"),
          cl_arg = c("-of", "COG", "-co", "BLOCKSIZE=512", "-co", "SPARSE_OK=YES",
                     "-co", "COMPRESS=DEFLATE", "-co", "OVERVIEW_RESAMPLING=NEAREST",
                     "-co", "NUM_THREADS=ALL_CPUS"))
```

Things I have not done yet, and want to:

- Time this step, so there is an honest end-to-end "how long to get a COG" number
  against the original serial loop.
- Check whether reading a sparse Zarr into the COG driver preserves sparsity as well
  as writing blocks directly did (`SPARSE_OK=YES` should skip the absent chunks, which
  come back as nodata).
- The Zarr driver has no classic-API overviews; for a Zarr-native product the
  pyramid would be additional arrays (the multiscales convention in GDAL 3.13 for
  Zarr V3), which is a separate post.

## Appendix: populating without GDAL in the workers

The other question I set out with was whether GDAL could instantiate the store and
something more generic could fill it. It can. A Zarr V2 chunk is: the full chunk shape
(edge chunks padded with `fill_value`), C order, the declared dtype (`<i4` here),
passed through the declared compressor, saved as `row.col` in the array directory,
0-based. If the store is created with `COMPRESS=ZLIB`, base R already has the encoder,
because `memCompress(type = "gzip")` emits a zlib stream, which is what the numcodecs
`zlib` codec expects.

```{r}
write_chunk_raw <- function(mat, row_chunk, col_chunk, array_dir, chunk_size) {
  full <- matrix(0L, chunk_size, chunk_size)
  full[seq_len(nrow(mat)), seq_len(ncol(mat))] <- mat
  bytes <- writeBin(as.integer(t(full)), raw(), size = 4L, endian = "little")
  writeBin(memCompress(bytes, type = "gzip"),
           file.path(array_dir, sprintf("%d.%d", row_chunk, col_chunk)))
}
```

This has no shared writes of any kind, not even the metadata files, and it is the
mirror image of a Zarr *reader*, which is where I am headed with `zaro`. Open
questions to explore:

- ZSTD or Blosc from R without pulling in a heavy dependency, since ZLIB is the slow
  codec.
- Whether the worker should write chunks at all, or emit `(chunk key, bytes)` pairs so
  the sink can be a local directory, an object store, or something that builds a
  kerchunk/Icechunk reference set instead.
- Doing the same for the coverage-mode burn (`Float32`, `<f4`), where the chunks are
  no longer mostly a single value and the compressor choice matters more.

::: {.callout-note}
## A note on the surrounding tools

As before, this leans on tools in active development - `controlledburn`, `grout`,
`vaster` - and on `gdalraster` and `mirai`, which are not. The Zarr driver's behaviour
under concurrent update (metadata rewritten only when dirtied, empty chunks skipped)
is what I observed here and what a reading of the driver suggests, but it is not a
documented guarantee; check the `.zmetadata` mtime and the chunk-file count on a small
grid before trusting it on a large one.
:::
