---
title: "What's on geodata.ucdavis.edu and how to access it"
author: "Michael Sumner"
date: "2026-03-17"
categories:
  - code
  - news
  - raster
  - vector
  - gdal
  - spatial
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
```

The R package [geodata](https://CRAN.R-project.org/package=geodata) provides
functions for downloading climate, elevation, soil, crop, species occurrence,
and administrative boundary data. All of its raster and vector data is hosted
on a single Apache web server at `geodata.ucdavis.edu`.

Every file on that server is directly accessible via HTTP. This means we can
enumerate, inspect, and in some cases read them using GDAL's virtual
filesystem — without downloading anything first.

This post documents the full site structure, the file formats served, and
which files can be streamed via `/vsicurl/` versus which need to be or should (for efficiency)be downloaded.

## Helper functions

The code snippets below use a small set of helper functions that wrap
`gdalraster::vsi_read_dir()` and related GDAL calls. Source them with:
 
```{r}
source("https://github.com/mdsumner/quarto-blog/releases/download/latest/geodata-ucdavis-tools.R")
```

This provides: `vsi_list()`, `geodata_files()`, `gdalinfo()`,
`check_block_size()`, and `streamability_report()`.

## Site structure

The server has three top-level paths that contain spatial data:

| Path | Contents |
|:-----|:---------|
| `/climate/` | WorldClim 2.1 and 1.4, CMIP5 legacy data |
| `/cmip6/` | Downscaled CMIP6 projections |
| `/geodata/` | Soil, landcover, crops, elevation, travel time, OSM, buildings |
| `/gadm/` | GADM administrative boundaries (versions 2.8, 3.6, 4.0, 4.1) |

The root URL itself and some intermediate directories serve an `index.html`
that says "Hello World!" rather than an Apache directory listing. The data
directories below these do have standard Apache autoindex enabled, which is
what makes `vsi_read_dir()` work.

## Enumerating files

GDAL's `/vsicurl/` handler can parse Apache directory listings. The
`gdalraster` package exposes this via `vsi_read_dir()`:

```{r}
library(gdalraster)

## flat directory — returns filenames
vsi_read_dir("/vsicurl/https://geodata.ucdavis.edu/geodata/soil/soilgrids/")
```

For structured enumeration with URLs and GDAL dataset names:

```{r}
soilgrids <- geodata_files("geodata/soil/soilgrids")
head(soilgrids)
```

This returns a data.frame with columns `url`, `file`, `ext`, and `gdaldsn`.
The `gdaldsn` column contains the `/vsicurl/` path for files GDAL can read,
and `NA` for files it cannot (`.rds`, `.txt`, etc.).

## What's there: a complete inventory

### SoilGrids (`/geodata/soil/soilgrids/`)

Global soil properties at 30 arc-second resolution. 48 TIF files covering
8 properties (bdod, cec, cfvo, clay, nitrogen, ocd, phh2o, sand, silt, soc)
at 4 depth intervals, plus pH quantiles. Files range from 83 MB to 546 MB.

```{r}
soilgrids <- geodata_files("geodata/soil/soilgrids", pattern = "\\.tif$")
nrow(soilgrids)
```

**Format**: Single-band GeoTIFF, LZW compressed, `Block=43200x1`
(row-striped). No overviews. No internal tiling.

```{r}
check_block_size(soilgrids$gdaldsn[1])

```

**Streamability**: Row-striped layout means sub-region reads via `/vsicurl/`
require decompressing entire rows spanning the full 43200 pixel width.
Reading a small bounding box will transfer most of the file. These files are
best downloaded locally.

The `geodata::soil_world_vsi()` function constructs `/vsicurl/` paths to
these same files — the name suggests streaming is the intended use, but the
row-striped layout makes it impractical for anything less than a large
regional or global read.

### WorldClim 2.1 tiles (`/climate/worldclim/2_1/tiles/tile/`)

30 arc-second resolution climate data, pre-tiled into 72 spatial tiles.
8 variables (bio, elev, prec, srad, tavg, tmax, tmin, wind). Bio tiles
have 19 bands; monthly variables have 12 bands; elevation has 1. File sizes
range from ~5 MB (ocean/ice tiles) to ~550 MB (land-dense tiles).

```{r}
wc_tiles <- geodata_files("climate/worldclim/2_1/tiles/tile", pattern = "\\.tif$")
nrow(wc_tiles)  ## expect 576 = 72 tiles × 8 variables
```

**Format**: Multi-band GeoTIFF. Block layout should be checked:

```{r}
check_block_size(wc_tiles$gdaldsn[1])
```

### WorldClim 2.1 global (`/climate/worldclim/2_1/base/`)

Global grids at 10m, 5m, 2.5m, and 30s resolution. All stored as `.zip`
files containing one or more TIF files per variable.

| Resolution | Variables | Example size |
|:-----------|:----------|:-------------|
| 10 minute | bio, elev, prec, srad, tavg, tmax, tmin, vapr, wind | 1.3 MB – 48 MB |
| 5 minute | same | 4.6 MB – 171 MB |
| 2.5 minute | same | 17 MB – 628 MB |
| 30 second | same | 323 MB – 9.7 GB |

GDAL can read inside these zips via `/vsizip//vsicurl/`:

```{r}
## example: 10-minute bioclim, first file inside the zip
dsn <- "/vsizip//vsicurl/https://geodata.ucdavis.edu/climate/worldclim/2_1/base/wc2.1_10m_bio.zip"
## list the contents of the zip
vsi_read_dir(dsn)
```

The internal TIF filenames follow the pattern
`wc2.1_{res}_{var}_{number}.tif`.

### WorldClim 2.1 country extracts (`/climate/worldclim/2_1/tiles/iso/`)

30 arc-second data clipped to individual countries, identified by ISO 3166-1
alpha-3 codes.

```{r}
iso <- geodata_files("climate/worldclim/2_1/tiles/iso")
head(iso)
```

### WorldClim 1.4 (`/climate/worldclim/1_4/`)

Legacy WorldClim version in BIL (Band Interleaved by Line) format inside
zip files. Three subdirectories: `cur/` (current), `fut/` (future
projections), `pst/` (past / LGM / mid-Holocene).

BIL format is GDAL-readable but, like the row-striped TIFs, does not support
efficient random access.

### CMIP6 projections (`/cmip6/`)

Downscaled and bias-corrected CMIP6 future climate data. Organised as:

```
/cmip6/{resolution}/{model}/{ssp}/
```

Resolutions: `30s`, `2.5m`, `5m`, `10m`. 23 GCMs. 4 SSPs (ssp126, ssp245,
ssp370, ssp585). 4 time periods (2021-2040, 2041-2060, 2061-2080,
2081-2100). Variables: bioc, prec, tmax, tmin.

At 30s resolution these are bare TIF files, some exceeding 20 GB.

```{r}
## list models at 2.5 minute resolution
vsi_read_dir("/vsicurl/https://geodata.ucdavis.edu/cmip6/2.5m/")
```

There is also a `/cmip6/tiles/` directory with tiled versions.

### CMIP5 projections (`/climate/cmip5/`)

Legacy CMIP5 data at 30s, 2.5m, 5m, 10m resolutions. Separate directories
for `lgm/` (Last Glacial Maximum) and `mid/` (mid-Holocene).

### Elevation

TODO: locate and document. The `geodata` package provides `elevation_3s()`,
`elevation_30s()`, and `elevation_global()` — the URLs for these need to be
traced from the source code.

### Landcover (`/geodata/landcover/`)

Two subdirectories:

- `/geodata/landcover/30s/` — 30 arc-second aggregated landcover fractions
- `/geodata/landcover/3deg_cogs/` — ESA WorldCover 10m data as 3-degree
  tiles, labelled as COGs in the directory name

```{r}
lc_cogs <- geodata_files("geodata/landcover/3deg_cogs", pattern = "\\.tif$")
nrow(lc_cogs)
```

The `3deg_cogs` files follow the naming convention
`ESA_WorldCover_10m_2020_v100_{lat}{lon}_Map.tif`. If these are genuine
COGs (internally tiled with overviews), they would be efficiently
streamable via `/vsicurl/`.

```{r}
## check whether these are really COGs
check_block_size(lc_cogs$gdaldsn[1])

```

### Crops (`/geodata/crops/`)

Subdirectories for SPAM and Monfreda crop data, plus cropland density and
crop calendars. File formats and structure need to be enumerated.

### GADM administrative boundaries (`/gadm/gadm4.1/`)

Per-country boundaries in multiple formats:

| Subdirectory | Format | GDAL-readable | Notes |
|:-------------|:-------|:--------------|:------|
| `gpkg/` | GeoPackage | Yes | SQLite-based, supports HTTP range requests |
| `shp/` | Shapefile (zipped) | Yes, via /vsizip/ | Multiple component files |
| `json/` | GeoJSON | Yes | |
| `pck/` | RDS (R serialization) | No | Used by `geodata::gadm()` |
| `kmz/` | KMZ (zipped KML) | Yes, via /vsizip/ | |

Whole-dataset archives:

- `gadm_410-gpkg.zip` — 1.4 GB, all countries in one GeoPackage
- `gadm_410-gdb.zip` — 913 MB, FileGDB format
- `gadm_410-levels.zip` — 2.5 GB, separated by admin level
- `gadm_410-raw.gpkg` — 2.6 GB, bare GeoPackage on HTTP

The `gadm_410-raw.gpkg` is directly streamable:

```{r}
## open GADM as a single GeoPackage over HTTP
dsn <- "/vsicurl/https://geodata.ucdavis.edu/gadm/gadm4.1/gadm_410-raw.gpkg"
## list layers
gdalraster::ogr_ds_layer_names(dsn)
```

Per-country GeoPackage files are also streamable:

```{r}
dsn <- "/vsicurl/https://geodata.ucdavis.edu/gadm/gadm4.1/gpkg/gadm41_AUS.gpkg"
gdalraster::ogr_ds_layer_names(dsn)
```

### Other datasets

| Path | Contents | geodata function |
|:-----|:---------|:-----------------|
| `/geodata/travel/` | Travel time rasters | `travel_time()` |
| `/geodata/osm/` | OpenStreetMap extracts (highways, places) | `osm()` |
| `/geodata/buildings/` | Building footprint data | — |

These need further enumeration.

### Datasets not on this server

Some `geodata` functions fetch data from external APIs rather than
`geodata.ucdavis.edu`:

- `sp_occurrence()` — queries the GBIF API
- `bio_oracle()` — queries the Bio-Oracle server at `bio-oracle.org`

## Block layout and streamability

The central question for remote access is whether a file's internal layout
supports efficient HTTP range requests. GDAL's `/vsicurl/` handler issues
byte-range requests to read specific portions of a file. This works well
when the file is internally tiled (e.g., COG format with `Block=256x256`)
but poorly when the file is row-striped (`Block=width×1`).

For a row-striped file, reading a small spatial window requires
decompressing one or more full rows, each spanning the entire raster width.
For a 43200-pixel-wide global grid, every row is a complete scan line from
-180° to +180°. A single HTTP range request per row, times the number of
rows in the window, with full LZW decompression of each row. For a small
bounding box this is extremely wasteful.

### What we know

| Collection | Block layout | Streamable? |
|:-----------|:-------------|:------------|
| SoilGrids TIFs | 43200×1 (row-striped) | No — download |
| WorldClim 2.1 tiles | TODO: check | Probably row-striped |
| WorldClim 2.1 global zips | N/A (zip) | Via /vsizip/, but still row-striped inside |
| CMIP6 30s TIFs | TODO: check | Unknown |
| Landcover 3deg COGs | TODO: check | Likely yes (name says COGs) |
| Landcover 30s | TODO: check | Unknown |
| GADM GeoPackage | N/A (vector) | Yes — SQLite random access works over HTTP |
| GADM raw GeoPackage | N/A (vector) | Yes |

### How to check

```{r, eval=FALSE}
## for any TIF on the server
bs <- check_block_size("/vsicurl/https://geodata.ucdavis.edu/path/to/file.tif")
bs
## if is_row_striped is TRUE: download, don't stream
## if is_tiled is TRUE: streaming is efficient
```

## Generating file tables

For any collection, generate a complete table of URLs and GDAL dataset names:

```{r}
## SoilGrids: all TIFs
sg <- geodata_files("geodata/soil/soilgrids", pattern = "\\.tif$")
head(sg[, c("url", "gdaldsn")])
```

```{r}
## WorldClim tiles: all TIFs
wc <- geodata_files("climate/worldclim/2_1/tiles/tile", pattern = "\\.tif$")
head(wc[, c("url", "gdaldsn")])
```

```{r}
## GADM 4.1 GeoPackages
gadm <- geodata_files("gadm/gadm4.1/gpkg", pattern = "\\.gpkg$")
head(gadm[, c("url", "gdaldsn")])
```

```{r}
## CMIP6 models at 2.5 minute resolution
## first list the model directories, then enumerate within each
models <- vsi_read_dir("/vsicurl/https://geodata.ucdavis.edu/cmip6/2.5m/")
models
```

## Still to document

Several areas need further investigation:

- Elevation data paths (`elevation_3s`, `elevation_30s`, `elevation_global`)
- Exact contents of `/geodata/crops/`, `/geodata/travel/`, `/geodata/osm/`,
  `/geodata/buildings/`
- Block layout of WorldClim tiles, CMIP6 30s TIFs, and landcover COGs
- Contents of `/geodata/landcover/30s/`
- Contents of `/geodata/soil/soilgrids/agg/`, `fill/`, and `raw/`
  subdirectories
- The `/cmip6/tiles/` tiled variant structure

Once these are filled in, the tables generated by `geodata_files()` will
provide a complete machine-readable inventory of the site.


