---
title: "Distance, perimeter and area in sf and terra: which number do you get?"
author: "Michael Sumner"
date: "2026-09-24"
format:
  html:
    code-fold: false
    toc: true
execute:
  echo: true
  warning: true
  message: false
---

```{r setup}
#| include: false
options(digits = 9)
```

In May 2025 a question on R-sig-Geo asked why `sf::st_perimeter()` gave a
different answer from `sf::st_length(sf::st_boundary())` for the same small
land parcel in France, with spherical geometry switched off
([thread](https://stat.ethz.ch/pipermail/r-sig-geo/2025-May/029541.html)).
There were three different numbers for the perimeter of that parcel, and each
of them is correct for a particular model of the Earth and a particular
meaning of "edge". I posted a
[gist](https://gist.github.com/mdsumner/5daedfcbee851ca31a7aa9006dbf74cf)
at the time as a start at unpicking them. (In my message on that thread,
"sf+longlat-s2 uses Karney" should read "with s2 switched off".)

The specific inconsistency in `st_perimeter()` has since been fixed in sf
([#2541](https://github.com/r-spatial/sf/issues/2541)). What remains is the
more general question this post is about: for a given geometry, how does each
package decide which model of the Earth to measure on, and can a user predict
the answer from the data in front of them?

The short version:

* **terra** decides from the geometry's coordinate reference system. Longitude
  and latitude input is always measured on the ellipsoid. Projected input is
  measured on the ellipsoid for area and cell size (by default), and in the
  plane for perimeter and distance.
* **sf** decides from the coordinate reference system *and* a session option,
  `sf_use_s2()`. Longitude and latitude input is measured on a sphere when s2
  is on (the default) and on the ellipsoid when it is off. Projected input is
  always measured in the plane. The same option also selects the engine used
  for predicates and overlay operations.

The rest of the post establishes this with code, and then looks at how large
the differences are and where they come from.

## Versions

Everything below depends on these versions, and the results should be re-run
when they change.

```{r versions}
pkgs <- c("sf", "terra", "s2", "lwgeom", "geosphere", "geodist", "wk")
vapply(pkgs, function(p) as.character(utils::packageVersion(p)), "")
sf::sf_extSoftVersion()
terra::gdal(lib = "all")
```

## The parcel

This is the polygon from the thread, as a plain coordinate matrix in
Lambert-93 (EPSG:2154, RGF93 on the GRS 1980 ellipsoid). It is about 40 m
across. We keep three versions of it: projected, transformed to longitude and
latitude (EPSG:4326), and with no CRS at all.

```{r fixture}
xy <- matrix(c(
  320102.67, 6845954.56,
  320111.85, 6845962.15,
  320131.82, 6845971.55,
  320134.33, 6845971.90,
  320138.50, 6845953.21,
  320141.93, 6845936.12,
  320135.33, 6845935.07,
  320118.58, 6845931.54,
  320113.45, 6845929.91,
  320105.18, 6845950.57,
  320102.67, 6845954.56
), ncol = 2, byrow = TRUE)

p   <- sf::st_sfc(sf::st_polygon(list(xy)), crs = "EPSG:2154")  ## projected
pll <- sf::st_transform(p, "EPSG:4326")                          ## longlat
p0  <- sf::st_set_crs(p, NA)                                     ## no crs
ll  <- sf::st_coordinates(pll)[, 1:2]
seg_i <- seq_len(nrow(xy) - 1)
```

A small helper sets `sf_use_s2()` for one expression and restores it
afterwards.

```{r with-s2}
with_s2 <- function(use, code) {
  old <- suppressMessages(sf::sf_use_s2(use))
  on.exit(suppressMessages(sf::sf_use_s2(old)))
  force(code)
}
```

## Three reference numbers

Before calling any sf or terra function, we compute the perimeter and area
directly under each of the three models, so that every later result can be
labelled by which model it corresponds to.

* **Planar**: each edge is a straight line in the projected coordinates, and
  lengths and areas are Euclidean in those coordinates.
* **Ellipsoid**: each edge is a geodesic on the WGS84 ellipsoid, computed with
  Karney's algorithms (GeographicLib), here via geosphere.
* **Sphere**: each edge is a great circle on a sphere of radius 6371010 m,
  which is the radius s2 uses.

```{r references}
planar_perim <- sum(sqrt(rowSums((xy[seg_i + 1, ] - xy[seg_i, ])^2)))
planar_area  <- abs(sum(xy[seg_i, 1] * xy[seg_i + 1, 2] -
                        xy[seg_i + 1, 1] * xy[seg_i, 2])) / 2

karney_perim <- sum(geosphere::distGeo(ll[seg_i, ], ll[seg_i + 1, ]))
karney_area  <- geosphere::areaPolygon(ll)

R_s2 <- 6371010
sphere_perim <- s2::s2_perimeter(s2::as_s2_geography(pll), radius = R_s2)
sphere_area  <- s2::s2_area(s2::as_s2_geography(pll), radius = R_s2)

refs_perim <- c(planar = planar_perim, ellipsoid = karney_perim, sphere = sphere_perim)
refs_area  <- c(planar = planar_area,  ellipsoid = karney_area,  sphere = sphere_area)
refs_perim
refs_area

which_number <- function(x, refs) {
  hit <- names(refs)[abs(refs - x) / refs < 1e-6]
  if (length(hit) == 0) "OTHER" else paste(hit, collapse = "/")
}
```

The perimeter is 129.249 m planar, 129.285 m on the ellipsoid and 129.098 m
on the sphere. The area is 1049.65, 1050.24 and 1047.03 square metres
respectively.

### What each number is

The ellipsoidal length of an edge is the same as its planar length in an
azimuthal equidistant projection centred on one of its endpoints, because that
projection preserves distance from its centre.

```{r aeqd}
aeqd_edge <- function(a, b) {
  crs <- sprintf("+proj=aeqd +lon_0=%.12f +lat_0=%.12f +ellps=WGS84", a[1], a[2])
  q <- terra::project(rbind(a, b), from = "EPSG:4326", to = crs)
  sqrt(sum((q[2, ] - q[1, ])^2))
}
c(aeqd = sum(vapply(seg_i, function(i) aeqd_edge(ll[i, ], ll[i + 1, ]), 0)),
  karney = karney_perim)
```

The spherical number is the same geodesic calculation with zero flattening and
s2's radius.

```{r sphere-as-karney}
sum(geosphere::distGeo(ll[seg_i, ], ll[seg_i + 1, ], a = R_s2, f = 0))
```

The planar number is the ellipsoidal number scaled by the projection. Lambert-93
is conformal, so at a point it has a single scale factor k in every direction.
Lengths scale by k and areas by k squared. Measuring k numerically at the parcel
reproduces both ratios.

```{r scale-factor}
k_perim <- planar_perim / karney_perim
k_area  <- planar_area / karney_area

cen <- colMeans(ll[seg_i, ])
nb  <- geosphere::destPoint(cen, b = 45, d = 10)
cq  <- terra::project(rbind(cen, nb), from = "EPSG:4326", to = "EPSG:2154")
k_point <- sqrt(sum((cq[2, ] - cq[1, ])^2)) / 10

c(k_point = k_point, k_perim = k_perim, k_perim_sq = k_perim^2, k_area = k_area)
```

At this location k is 0.999718, so the planar perimeter is 0.028% shorter than
the ellipsoidal one and the planar area is 0.056% smaller. An equal-area
projection centred on the parcel gives the ellipsoidal area to within about
one part in 10^8.

```{r laea}
laea <- sprintf("+proj=laea +lon_0=%.12f +lat_0=%.12f +ellps=WGS84", cen[1], cen[2])
c(laea = as.numeric(sf::st_area(sf::st_set_crs(sf::st_transform(pll, laea), laea))),
  karney = karney_area)
```

## How each package decides

### sf

The decision in sf is visible in the source of `st_perimeter()`. The first
branch is taken when `sf_use_s2()` is `TRUE` and the geometry is longitude and
latitude.

```{r sf-body}
body(sf::st_perimeter)
```

The sf documentation for `st_area()` describes the same rule: with longitude
and latitude coordinates, area is computed on the ellipsoid with
`st_geod_area()` if `sf_use_s2()` is `FALSE`, and on the sphere with
`s2_area()` if it is `TRUE`; with projected coordinates, area is computed in
the plane. `st_length()` and `st_distance()` follow the same pattern.

When s2 is off, the longitude and latitude branch of `st_perimeter()` goes via
`st_length(st_boundary(x))`. It does not call lwgeom's perimeter, which
declines longitude and latitude input:

```{r lwgeom-perim}
tryCatch(lwgeom::st_perimeter(pll), error = function(e) conditionMessage(e))
```

### terra

terra has no global option for this. For longitude and latitude input, its
documentation states that distances, lengths and areas are computed on the
WGS84 ellipsoid with Karney's algorithms. For projected input, the behaviour
differs between functions. `expanse()` and `cellSize()` have an argument
`transform`, which defaults to `TRUE` and computes the result on the
ellipsoid. `perim()` and `distance()` have no such argument and work in the
plane.

```{r terra-audit}
vp  <- terra::vect(p)
pts <- terra::vect(xy[1:2, ], crs = "EPSG:2154")
r   <- terra::rast(terra::ext(320100, 320150, 6845925, 6845975), res = 10, crs = "EPSG:2154")
unlist(list(
  perim          = terra::perim(vp),
  expanse        = terra::expanse(vp),
  expanse_plane  = terra::expanse(vp, transform = FALSE),
  distance       = as.numeric(terra::distance(pts)),
  cellSize       = terra::global(terra::cellSize(r, unit = "m"), "mean")[1, 1],
  cellSize_plane = terra::global(terra::cellSize(r, unit = "m", transform = FALSE), "mean")[1, 1]
))
## references for the distance row
c(planar = sqrt(sum((xy[2, ] - xy[1, ])^2)), karney = geosphere::distGeo(ll[1, ], ll[2, ]))
```

The area of a nominal 10 m by 10 m cell is 100.056 square metres on the
ellipsoid, which is 100 divided by k squared. The distance between the first
two vertices is the planar value, 11.911 m, rather than the ellipsoidal
11.915 m.

## The grid

Now every combination: four ways of asking for a perimeter, three ways of
asking for an area, three versions of the polygon, and `sf_use_s2()` on and
off. Each result is labelled by the reference number it matches.

```{r grid-setup}
inputs <- list(projected = p, longlat = pll, none = p0)

perim_calls <- list(
  "sf::st_perimeter"        = function(x) sf::st_perimeter(x),
  "sf::st_length(boundary)" = function(x) sf::st_length(sf::st_boundary(x)),
  "sf::st_length(cast MLS)" = function(x) sf::st_length(sf::st_cast(x, "MULTILINESTRING")),
  "terra::perim"            = function(x) terra::perim(terra::vect(x))
)
area_calls <- list(
  "sf::st_area"                     = function(x) sf::st_area(x),
  "terra::expanse"                  = function(x) terra::expanse(terra::vect(x)),
  "terra::expanse(transform=FALSE)" = function(x) terra::expanse(terra::vect(x), transform = FALSE)
)

run_grid <- function(calls, refs) {
  out <- expand.grid(call = names(calls), input = names(inputs), s2 = c(TRUE, FALSE),
                     stringsAsFactors = FALSE)
  out$value <- NA_real_
  out$msg <- ""
  for (i in seq_len(nrow(out))) {
    f <- calls[[out$call[i]]]
    x <- inputs[[out$input[i]]]
    r <- tryCatch(
      withCallingHandlers(
        with_s2(out$s2[i], as.numeric(f(x))),
        warning = function(w) {
          out$msg[i] <<- conditionMessage(w)
          invokeRestart("muffleWarning")
        }),
      error = function(e) {
        out$msg[i] <<- paste("ERROR:", conditionMessage(e))
        NA_real_
      })
    out$value[i] <- r
  }
  out$matches <- vapply(out$value, function(v) if (is.na(v)) "-" else which_number(v, refs), "")
  out
}

wide <- function(g) {
  w <- reshape(g[, c("call", "input", "s2", "matches")],
               idvar = c("call", "s2"), timevar = "input", direction = "wide")
  names(w) <- sub("^matches\\.", "", names(w))
  rownames(w) <- NULL
  w
}
```

```{r grid-perim}
perim_grid <- run_grid(perim_calls, refs_perim)
wide(perim_grid)
```

```{r grid-area}
area_grid <- run_grid(area_calls, refs_area)
wide(area_grid)
```

The three sf routes to a perimeter now agree with each other in every case,
which was not true at the time of the original thread. With projected input,
or no CRS, everything in sf is planar. With longitude and latitude input, sf
gives the spherical number when s2 is on and the ellipsoidal number when it is
off.

terra does not read `sf_use_s2()`, so its rows are the same in both halves of
each table. With longitude and latitude input it gives the ellipsoidal number
for both perimeter and area. With projected input, `perim()` is planar and
`expanse()` is ellipsoidal. With no CRS, terra computes in the plane and warns
that the results may be wrong.

For longitude and latitude data, which is the common case for data arriving
from files and web services, the two packages with their default settings
give different answers: sf gives 129.098 m and 1047.03 square metres, terra
gives 129.285 m and 1050.24 square metres.

## One option, two roles

In sf, `sf_use_s2()` selects the model for measures and also the engine for
predicates and overlays. With s2 off, longitude and latitude geometries go to
GEOS, which treats the coordinates as planar. Measures, however, go to
GeographicLib on the ellipsoid, where each edge is a geodesic. So within one
session with s2 off, the same polygon has ellipsoidal geodesic edges for the
purpose of its area and straight edges in longitude and latitude for the
purpose of point-in-polygon.

This matters for wide polygons at high latitude. A geodesic between two points
at the same latitude does not follow the parallel; it bends toward the pole.
Here is a box 120 degrees wide between 70 and 60 degrees south, and a point
just inside its northern edge.

```{r coupling}
bx <- sf::st_sfc(sf::st_polygon(list(rbind(
  c(-60, -70), c(60, -70), c(60, -60), c(-60, -60), c(-60, -70)))),
  crs = "EPSG:4326")
pt <- sf::st_sfc(sf::st_point(c(0, -62)), crs = "EPSG:4326")

with_s2(TRUE,  list(area = sf::st_area(bx),
                    intersects = sf::st_intersects(bx, pt, sparse = FALSE)[1, 1]))
with_s2(FALSE, list(area = sf::st_area(bx),
                    intersects = sf::st_intersects(bx, pt, sparse = FALSE)[1, 1]))
```

With s2 on, the area is spherical and the point is outside the polygon, which
is consistent: on the sphere the northern edge passes south of 62 degrees
south at longitude 0. With s2 off, the area is ellipsoidal, which is the more
accurate model, but the point is now inside the polygon, because GEOS treats
the northern edge as the straight line at 60 degrees south. A user who
switches s2 off to obtain ellipsoidal measures therefore also changes the
answers to spatial predicates, and the measures and predicates no longer
describe the same polygon.

## Point distance and two spheres

For a single pair of points (the first two vertices of the parcel) there is a
fourth number.

```{r point-distance}
a <- ll[1, , drop = FALSE]
b <- ll[2, , drop = FALSE]
pa <- sf::st_sfc(sf::st_point(a), crs = "EPSG:4326")
pb <- sf::st_sfc(sf::st_point(b), crs = "EPSG:4326")
c(
  sf_s2     = with_s2(TRUE,  as.numeric(sf::st_distance(pa, pb))),
  sf_nos2   = with_s2(FALSE, as.numeric(sf::st_distance(pa, pb))),
  terra_geo = terra::distance(a, b, lonlat = TRUE),
  terra_hav = terra::distance(a, b, lonlat = TRUE, method = "haversine"),
  geodist   = geodist::geodist(a, b, measure = "geodesic"),
  karney    = geosphere::distGeo(a, b),
  planar    = sqrt(sum((xy[2, ] - xy[1, ])^2))
)
```

sf with s2 off, terra's default method, geodist and geosphere all agree on the
ellipsoidal distance. sf with s2 on gives the spherical distance. terra's
`method = "haversine"` gives a third value, because it uses a sphere with the
equatorial radius of WGS84 rather than s2's mean radius:

```{r haversine-radius}
c(terra_hav      = terra::distance(a, b, lonlat = TRUE, method = "haversine"),
  hav_equatorial = geosphere::distHaversine(a, b, r = 6378137),
  hav_mean_s2    = geosphere::distHaversine(a, b, r = R_s2))
```

The two spheres differ in radius by a factor of 1.0011, so "spherical
distance" is itself not one number.

<!-- TODO check at terra tip: the SpatRaster distance() method appeared to
default to method = "haversine" in the CRAN source I read. Confirm with
  getMethod("distance", signature(x = "SpatRaster", y = "missing"))
and mention here if it still does. -->

## How large is the difference, and where?

For a 40 m parcel the differences are centimetres. It is tempting to conclude
that the choice of model only matters for large polygons. For the sphere
against the ellipsoid, that is not the case: the relative error in area is set
by latitude and is essentially the same for a box one metre across and a box
one degree across.

```{r sweep}
box <- function(lon0, lat0, size) {
  m <- rbind(c(lon0, lat0), c(lon0 + size, lat0), c(lon0 + size, lat0 + size),
             c(lon0, lat0 + size), c(lon0, lat0))
  sf::st_sfc(sf::st_polygon(list(m)), crs = "EPSG:4326")
}
sweep_lat <- function(size, lats = seq(-89.5, 89.5 - size, by = 1)) {
  do.call(rbind, lapply(lats, function(lat0) {
    g <- box(147, lat0, size)
    k <- geosphere::areaPolygon(sf::st_coordinates(g)[, 1:2])
    data.frame(lat = lat0 + size / 2, size = size,
               s2 = with_s2(TRUE, as.numeric(sf::st_area(g))) / k - 1,
               terra = terra::expanse(terra::vect(g)) / k - 1)
  }))
}
sw <- rbind(sweep_lat(0.001), sweep_lat(0.1), sweep_lat(1))
range(sw$terra)
```

```{r sweep-plot}
#| fig-cap: "Relative error of s2 spherical area against Karney ellipsoidal area, by latitude, for square boxes of three sizes. The three curves coincide."
plot(sw$lat, 100 * sw$s2, type = "n", xlab = "latitude",
     ylab = "area error vs ellipsoid (%)", main = "s2 sphere vs Karney ellipsoid")
sizes <- unique(sw$size)
for (s in sizes) {
  d <- sw[sw$size == s, ]
  lines(d$lat, 100 * d$s2, lty = match(s, sizes))
}
abline(h = 0, col = "grey")
legend("topright", legend = paste(sizes, "deg box"), lty = seq_along(sizes))
```

The spherical area is 0.45% too large at the equator, correct near 35 degrees
north and south, and 0.89% too small near the poles. At the latitudes of the
Southern Ocean and Antarctica the spherical area is consistently between 0.5%
and 0.9% smaller than the ellipsoidal area, regardless of the size of the
polygon. terra agrees with Karney to within floating point precision
throughout, since it uses the same library.

Planar error behaves differently: it depends on the projection's scale factor
over the polygon, which for a well-chosen local projection is small for small
polygons and grows with distance from the projection's centre or standard
lines.

## What is an edge?

All of the above assumes that everyone agrees on what the polygon is. The
coordinates alone do not determine that. Between two vertices, an edge could
be a geodesic on the ellipsoid, a great circle on the sphere, a straight line
in some projection, or a line of constant latitude. For small polygons these
coincide closely. For large ones they do not.

Here is a box 10 degrees on each side at 60 degrees south, first with only its
four corners, then densified at 0.1 degree spacing along its edges in
longitude and latitude so that the northern and southern edges follow the
parallels.

```{r edges}
b1  <- box(147, -60, 10)
b1d <- sf::st_set_crs(sf::st_segmentize(sf::st_set_crs(b1, NA), 0.1), "EPSG:4326")
c(
  corners_karney   = geosphere::areaPolygon(sf::st_coordinates(b1)[, 1:2]),
  densified_karney = geosphere::areaPolygon(sf::st_coordinates(b1d)[, 1:2]),
  corners_terra    = terra::expanse(terra::vect(b1)),
  densified_terra  = terra::expanse(terra::vect(b1d))
)
```

The two versions differ in area by 0.25%, which is of the same order as the
difference between the sphere and the ellipsoid at this latitude. terra and
Karney agree on each version, so both treat the edges as geodesics and neither
densifies implicitly.

The same question applies to terra's `transform = TRUE` for projected
polygons. The vertices are transformed to longitude and latitude and the edges
are then treated as geodesics. An edge that is straight in the projection is
not a geodesic, so for large polygons the result is the ellipsoidal area of a
slightly different polygon. Densifying in the projected coordinates before
measuring removes this.

## Summary

| input | sf, s2 on (default) | sf, s2 off | terra |
|---|---|---|---|
| longitude/latitude, length and perimeter | sphere | ellipsoid | ellipsoid |
| longitude/latitude, area | sphere | ellipsoid | ellipsoid |
| projected, length and perimeter | planar | planar | planar |
| projected, area | planar | planar | ellipsoid (`transform = FALSE` for planar) |
| no CRS | planar | planar | planar, with a warning |

For longitude and latitude data, terra's rule can be stated in one line: it
always measures on the ellipsoid. sf's answer depends on an option that is set
per session, is on by default, and also determines how predicates and
overlays are computed. The consequence is that an sf user who wants
ellipsoidal measures has to change the behaviour of their spatial predicates
as well, and a user who leaves the default gets a spherical measure whose
error varies from +0.45% to -0.89% with latitude, independent of the size of
the feature.

Two changes would make each package's rule simpler to state. In sf, a way to
select the ellipsoid for measures independently of the engine used for
predicates, for example an argument to the measure functions or a separate
option. In terra, extending the `transform` argument to `perim()` and
`distance()`, so that projected input is measured on the ellipsoid for every
quantity whenever the CRS is known, as it already is for area.

In the meantime, the grid above can be run against any installed versions,
and is a quick way to check which number a given function returns.
