---
title: "Rasterizing ice-free Antarctica"
author: "Michael Sumner"
date: "2026-08-19"
categories: [controlledburn, rasterization, Antarctica, GDAL]
description: >
  Sparse rasterization of 25,954 Antarctic rock outcrop polygons onto a
  31-billion-cell grid — in under a second.
execute:
  collapse: true
  comment: "#>"
fig-width: 8
fig-height: 8
format:
  html:
    code-fold: false
---

```{r}
#| label: setup
#| include: false
Sys.setenv("GDAL_DISABLE_READDIR_ON_OPEN" = "EMPTY_DIR")
```

Antarctica's exposed rock — the ice-free ground — has been mapped as
25,954 polygons covering nunataks, coastal outcrops, and mountain
ranges across the continent. The Reference Elevation Model of
Antarctica (REMA) provides a 2m DEM mosaic of the entire ice sheet,
tiled at multiple resolutions.

What does it take to rasterize those polygons onto the REMA grid? At
32m resolution, the grid is 170318 x 182568 pixels. A raster at that size would require enormous
memory, but controlledburn does it in under a second, producing 30-150 MB of sparse
output, then serves up any individual tile on demand.

::: {.callout-note collapse="false"}
## A note on the surrounding tools

This post reaches for `vapour`, `terra`, `gdalraster`, `grout`, and a few
others alongside `controlledburn` itself, without much ceremony about why
one is used here and another there. That's a fair reflection of where
things stand: `controlledburn` is under heavy, active development, and the
tooling around it — how you get a grid spec, read geometry, or write a
COG — has mattered less than getting the core right. Some of this
will likely consolidate as the package settles. Treat the package mix here
as a snapshot of current convenience, not a considered API.
:::

## Setup

Three cloud-hosted datasets, read directly via GDAL virtual filesystem
paths.

```{r}
#| label: data
# REMA 32m DEM mosaic — we read only the grid spec, no pixel values
dsn <- paste0(
  "/vsicurl/https://raw.githubusercontent.com/mdsumner/rema-ovr/",
  "main/rema-vrt/32m_dem_tiles.vrt")

# Antarctic rock outcrop polygons (v7.3, Burton-Johnson et al.)
shp <- paste0(
  "/vsizip/{/vsicurl/https://github.com/AustralianAntarcticDivision/",
  "rema.proc/raw/refs/heads/master/01_rock_classification/",
  "Medium_resolution_vector_polygons_of_Antarctic_rock_outcrop_-_",
  "VERSION_7.3.zip}/Medium resolution vector polygons of Antarctic ",
  "rock outcrop - VERSION 7.3/",
  "add_rock_outcrop_medium_res_polygon_v7.3.gpkg")

# REMA tile index — for choosing a view window
rema_tiles <- paste0(
  "/vsizip/{/vsicurl/https://data.pgc.umn.edu/elev/dem/setsm/REMA/",
  "indexes/REMA_Mosaic_Index_latest_shp.zip}/REMA_Mosaic_Index_v2_shp")
```

## Read

The raster is never opened — `vapour_raster_info()` reads only the
metadata (extent, dimensions, CRS). The polygons are read as raw WKB
bytes.

```{r}
#| label: read
library(terra)
tiles <- terra::vect(rema_tiles)

raster_info <- vapour::vapour_raster_info(dsn)
rock_info <- vapour::vapour_layer_info(shp)
rock <- wk::wkb(vapour::vapour_read_geometry(shp),
                 crs = rock_info$projection$Wkt)

cat(sprintf("Grid: %s × %s pixels (%.0f billion cells)\n",
    format(raster_info$dimension[1], big.mark = ","),
    format(raster_info$dimension[2], big.mark = ","),
    prod(as.numeric(raster_info$dimension)) / 1e9))
cat(sprintf("Polygons: %s geometries\n",
    format(length(rock), big.mark = ",")))
```

## Rasterize the continent

One call rasterizes every polygon onto the full continental grid.
The output is a sparse table of run-length encoded interior cells —
no dense matrix is allocated, no pixel values are touched.

```{r}
#| label: burn
library(controlledburn)

system.time({
  ba <- burn(rock, extent = raster_info$extent,
             dimension = raster_info$dimension, mode = "approx")
})

ba
cat(sprintf("\nObject size: %.0f MB\n", as.numeric(lobstr::obj_size(ba)) / 1e6))
```

Coverage mode computes exact boundary fractions — slower, but the total
burned area equals the true polygon area to floating-point precision:

```{r}
#| label: burn-coverage
system.time({
  be <- burn(rock, extent = raster_info$extent,
             dimension = raster_info$dimension, mode = "coverage")
})

be
```

## Look at a tile

The sparse output covers the whole continent but we can extract any
sub-window instantly with `crop_burn()`. Here: a 50 km REMA tile of the Windmill Islands, near Casey station.

```{r}
#| label: tile-locate
ll <- cbind(lon = 110.3, lat = -66.3)
xy <- reproj::reproj_xy(ll, crs(tiles), source = "EPSG:4326")
idx <- which(terra::is.related(tiles, vect(xy), "intersects"))
tile_ex <- as.vector(ext(tiles[idx, ]))
```

```{r}
#| label: tile-approx
#| fig-cap: "Rock outcrop near Windmill Islands, approx mode (cell-centre rule). White = exposed rock, dark = ice or ocean. Pink outlines are the source polygon boundaries."
ma <- materialize_chunk(crop_burn(ba, tile_ex))
cat(sprintf("Tile: %d × %d pixels\n", ncol(ma), nrow(ma)))

# Rock polygons that overlap this tile
rock_idx <- unlist(
  geos::geos_strtree_query(
    geos::geos_strtree(geos::as_geos_geometry(rock)),
    tiles[idx, ]))

par(bg = "#1a1a2e", mar = c(3, 3, 2, 1))
plot(tiles[idx, ], col = NA, border = NA,
     main = "Windmill Islands rock outcrop — 32m",
     col.main = "white", col.axis = "grey60")
ximage::ximage(ma, tile_ex, add = TRUE,
               col = c("#1a1a2e", "#e8d5b7"))

plot(rock[rock_idx], add = TRUE, border = "#e05780", lwd = 0.6)
```

```{r}
#| label: tile-coverage
#| fig-cap: "Same view, coverage mode. Boundary cells carry exact fractional weights — visible as soft edges where the polygon boundary crosses pixel boundaries."
mc <- materialize_chunk(crop_burn(be, tile_ex))

par(bg = "#1a1a2e", mar = c(3, 3, 2, 1))
plot(tiles[idx, ], col = NA, border = NA,
     main = "Windmill Islands rock outcrop — 32m, exact coverage",
     col.main = "white", col.axis = "grey60")
ximage::ximage(mc, tile_ex, add = TRUE,
               col = hcl.colors(64, "Lajolla", rev = TRUE))
plot(rock[rock_idx], add = TRUE, border = "#e05780", lwd = 0.6)
```

At 32m resolution the two modes are nearly identical. The difference
is sub-pixel boundary treatment:

```{r}
#| label: diff
n_diff <- sum(ma != mc)
cat(sprintf("Cells that differ: %s out of %s (%.3f%%)\n",
    format(n_diff, big.mark = ","),
    format(length(ma), big.mark = ","),
    100 * n_diff / length(ma)))
```

## Summary of the lazy rasterization process

controlledburn burned 25,954 polygons onto a 31-billion-cell grid in
under a second, producing a sparse run-length table small enough to
hold in memory. Any tile can be extracted and materialized on demand.
The raster was never read — only its grid specification. The polygons
were never validated or reprojected — just raw WKB bytes from a
cloud-hosted GeoPackage.

This is rasterization as an O(perimeter) operation: the cost scales
with the complexity of the polygon boundaries, not the number of
pixels in the grid. The sparse output is a reusable intermediate —
downstream processing (zonal statistics, masking, tiling) operates on
the same table without re-burning.

## Creating a raster file

Now we have a very dense version of our rock layer as a latent raster, we can
materialize arbitrary windows of it and write them to an actual raster file.
The strategy: create a SPARSE_OK Cloud-Optimized GeoTIFF (COG), iterate over
spatial tiles, crop the sparse burn result to each tile, materialize the dense
matrix, and write it via `gdalraster`. Tiles with no burned pixels are skipped
entirely — SPARSE_OK means unwritten tiles cost nothing on disk.

### Why `crop_burn()`?

`crop_burn()` is the spatial subsetting workhorse. Given a `controlledburn`
object covering an arbitrarily large grid, it filters the sparse tables (runs,
edges, lines, points) to a target extent and re-bases row/col indices to 1.
No dense matrix is ever allocated — it is pure data-frame filtering, so the
cost depends only on how many run-length entries fall inside the window, not
on how many pixels the full grid contains. This is what makes it feasible to
iterate over thousands of tiles without blowing up memory: each iteration
crops, materializes a small dense block, writes it, and discards it.

### Why super-tiles instead of native 512×512 blocks?

The COG internal tiling is 512×512, but writing one 512×512 block at a time
means ~120,000 write calls for the full grid. GDAL's `RasterIO()` accepts
arbitrary windows — they don't have to align with the internal block grid —
but it works best when the window is a multiple of the native block size.
Grouping into "super-tiles" of 2048×2048 (4×4 native blocks) brings the
tile count down to ~8,000 while keeping each dense allocation under 16 MB.
The dense output from `materialize_chunk()` is *not* itself tiled; it is a
plain row-major matrix. The tiling structure exists only in the GeoTIFF
container and is handled entirely by GDAL on write.

### The write loop

```{r}
#| label: geotiff-write
#| eval: false
library(controlledburn)
library(gdalraster)

outfile <- file.path(tempdir(), "rema_approx.tif")
block_size <- 512L

ext <- ba$extent
dm  <- ba$dimension
gdalraster::set_config_option("GDAL_NUM_THREADS", "ALL_CPUS")

ds <- create(
  format       = "COG",
  dst_filename = outfile,
  xsize = dm[1], ysize = dm[2],
  nbands   = 1,
  dataType = "Int32",
  options  = c(paste0("BLOCKSIZE=", block_size),
               "SPARSE_OK=YES",
               "COMPRESS=DEFLATE",
               "OVERVIEW_RESAMPLING=NEAREST",
               "NUM_THREADS=ALL_CPUS"),
  return_obj = TRUE
)

ds$setGeoTransform(vaster::extent_dim_to_gt(ext, dm))
ds$setNoDataValue(band = 1, 0)
```

Build a tile index with `grout`, then pre-filter to rows that contain any
burned data. This avoids even *visiting* tiles over open ocean or interior
ice sheet:

```{r}
#| label: geotiff-tiles
#| eval: false
superblock <- block_size * 4
gg    <- grout::grout(dm, ext, blocksize = superblock)
tiles <- grout::tile_index(gg)

# Keep only tile-rows that intersect at least one data row
data_rows <- unique(c(ba$runs$row, ba$edges$row,
                      ba$lines$row, ba$points$row))
tiles <- tiles[tiles$tile_row %in%
                 unique(ceiling(data_rows / superblock)), ]
```

The loop itself is straightforward: crop → materialize → write.

```{r}
#| label: geotiff-loop
#| eval: false
tiles_written <- 0L
for (i in seq_len(nrow(tiles))) {
  sub <- crop_burn(ba, c(tiles$xmin[i], tiles$xmax[i],
                         tiles$ymin[i], tiles$ymax[i]))

  if (nrow(sub$runs) == 0 && nrow(sub$edges) == 0) next

  mat <- materialize_chunk(sub, fun = "id")
  mat[is.na(mat)] <- 0

  ds$write(band = 1,
           xoff = tiles$offset_x[i], yoff = tiles$offset_y[i],
           xsize = tiles$ncol[i], ysize = tiles$nrow[i],
           rasterData = as.integer(t(mat)))

  tiles_written <- tiles_written + 1L
}

ds$flushCache()
ds$close()
```

```
## ~445 seconds elapsed
## tiles written: 3,482 of 7,511
## file size: 26,782 KB
## output: 170318 × 182568 Int32, nodata=0
```

The result is a valid COG — internally tiled, compressed, with overviews —
that weighs in at around 26 MB on disk for a 31-billion-cell grid. Only the
tiles that actually contain rock outcrop carry any data.

The compressed file is informationally equivalent to the sparse burn object
itself — both encode exactly the same set of non-empty cells. The burn
object in memory is `lobstr::obj_size(ba)` → 27.27 MB; the COG on disk is
26 MB. The sparse representation *is* the information content of the raster;
the file is just a different container for the same signal.

## Extracting point values

Not every downstream operation needs a dense raster. `extract_burn()` queries
the sparse tables directly at arbitrary point locations, with no
materialisation step at all. It matches interior runs by row equality and
column interval containment, and edges/lines/points by exact (row, col)
match. Cost is O(points + hits), independent of grid dimension.


```{r}
#| label: extract-points
#| eval: true
# Some coordinates in the burn's CRS (Antarctic Polar Stereographic)
pts <- structure(c(2412867, 2426358, 2424194, 2437557, 2437175, 2441121, 
2443539, 2443284, -911354, -909445, -915172, -913263, -913136, 
-912245, -914281, -920645), dim = c(8L, 2L))

plot(pts, asp = 1)
ximage::ximage(vapour::gdal_raster_nara(sds::wms_arcgis_mapserver_ESRI.WorldImagery_tms(), target_ext = tile_ex, target_crs = raster_info$projection, target_dim = c(512, 512)),add = TRUE)
points(pts, col = "green", pch = "+", cex = 2)
plot(rock[rock_idx], add = TRUE, border = "#e05780", lwd = 1.6)

e_approx <- extract_burn(ba, pts, fun = "id")

e_cover <- extract_burn(be, pts, fun = "sum")
cbind(approx = e_approx, cover = e_cover)
```

The function returns `NA` for background cells (no geometry touched that
pixel) and the geometry ID for cells inside a polygon. With `fun = "sum"`,
coverage-mode burns return exact fractional coverage at each point.  

This pattern generalizes. Because the sparse representation is a reusable
intermediate, other raster operations — zonal statistics, reclassification,
distance transforms — can be built on top of the same sparse tables without
re-burning or allocating a full dense grid. The burn is computed once;
everything else is derived.

## Packing, compression, and sparsity

A useful comparison: the Australian Antarctic Data Centre publishes
[Ice-free Antarctica](https://doi.org/10.26179/7mnh-j215) (Toth &
Terauds 2023) — a union of rock outcrop layers from imagery spanning
1960–2020, encoded as a single-band GeoTIFF at 100 m resolution
(52765 × 48132 pixels). Values are 0 (ice/ocean), 1, 2, or 3
(rock, by source layer), stored at 2 bits per sample (NBITS=2) in
128 × 128 tiles. The file is 613 MB.

That 2-bit packing is already a 4× saving over a byte raster —
but it is a fixed per-pixel economy. Every tile is materialized:
ocean, ice sheet, and rock alike each cost their full 4096 bytes.
The encoding is indifferent to the fact that the vast majority of
pixels say the same thing: zero. Recompressing the same tiles
with DEFLATE collapses the file to about 5.6 MB — a ~100×
reduction with zero information loss, because constant regions
become long runs of identical bytes, and long runs are the best
possible input to any general-purpose codec.

Our controlledburn COG encodes different (but related) data at
32 m resolution — a grid 3x denser in each dimension for 26 MB on disk. The sparse burn object in memory is
27 MB. These numbers are not directly comparable (different
polygons, different resolution, different pixel semantics), but
the *relationship* between them is the point: the information
content of "where is the rock?" is small relative to the grid
that contains it, regardless of how you choose to encode it.

The three approaches — bit-packing, codec compression, and
run-length sparse representation — are related lenses on the
same underlying reality. Bit-packing reduces the cost per pixel
but remains grid-dense. Compression recovers structure after the
fact, collapsing constant regions that the format didn't know to
avoid. Sparse rasterization exploits the same structure *before*
pixels exist: polygon interiors are runs by construction, so the
compact representation is produced directly rather than recovered
by a codec.

None of these is categorically superior. The AADC file is a
finished product, designed for broad interoperability and direct
random-access reads — the choice not to compress may well
reflect downstream tooling constraints or the simplicity of a
uniform tile stride. controlledburn's sparse format is an
intermediate: it is meant to sit between the vector source and
whatever raster product or analysis comes next. The value is in
seeing them as points on the same continuum — rasterization,
compression, and sparse encoding are all ways of negotiating
between the geometry of the data and the geometry of the grid.
